GPU-based Adaptive Beamforming for Active Sonar Imaging

### 2012-05-07 ECUA2012/gpu_capon_ecua2012.tex
la2pdf commit - 2012:05:07	23:56:40	(Mon 7.May)

Because each pixel in the image can be processed independently, the \gls{MVDR} beamformer is well suited for parallel acceleration. Of particular interest is the \gls{GPU}~\cite{Nilsen2009}, which is tailored for image processing, are available off-the-shelf at reasonable prices, and are found in most desktop computers already. A modern \gls{GPU} is a hardware architecture comprised of several hundred computing cores, each running a thread that executes a copy of a common function called a kernel. However, there are some inherent drawbacks with such a design: Each core is left with a rather limited amount of local memory, and distributing data to the computing cores and collecting the result is a processes with some inherent latency.

@@

To test the runtime speed of our \gls{GPU} implementation of \gls{MVDR}, we have compared it to an optimized single thread C and Matlab implementation (\Fig{benchmarks}). For this test we used a 55\,kpixels sectorscan image. The test system was a quad-core Intel Xeon E5620, with 64Gb of \gls{RAM}, and an nVidia Quadro 6000 card. This is a \gls{GPU} with 448 \gls{CUDA} cores and 6Gb of onboard \gls{RAM}, capable of roughly 1Tflops.

@@

As Figure \ref{benchmarks} illustrates, using a \gls{GPU} improves matters. If we look at the total runtimes for a subarray size of $L=16$ and $2N_\text{avg}+1=5$ pixels of temporal averaging, the \gls{GPU} implementation is able to process the test image in roughly 40\,ms.  For the same scenario the C implementation clocked in at roughly 3\,s and Matlab at 17\,s, which corresponds to a speedup factor of 75 and 435, respectively.

Interesting to note is also that solving $\eRi\1$ remain the main bottleneck in the \gls{GPU} implementation, and for larger $L$'s it dominates. However, this is a general Gauss Jordan solver which does not exploit symmetry properties of $\eR$, so creating e.g. a batch based Cholesky or UHDU solver should cut the runtimes quite a bit further. Both the C and Matlab implementation in this benchmark take this symmetry into account, which explains why the complexity curve of these implementation is roughly proportional with $L$, while the \gls{GPU} implementation is not.

HUGIN: 1.8m/s - range 250m \\
fs = 100kHz*30\%*$\frac{4}{3}$ = 40kHz \\
$N_{range px} = \frac{2\,250m}{1500m/s}40kHz = 13.3kpx$ \\
$\times 32$ beams = 426kpx \\\\

PRF = $\frac{1500m/s}{2 250m} = 3Hz$ \\
TP = 215px 3Hz = 1.28Mpx/s \\
4 arrays: 1.28Mpx/s * 4 = 5.12Mpx/s

\begin{itemize}
\item Need for speed: HUGIN 4 banks of 32 elements, can be processed faster than the ping repetition rate, with margins to spare.
\end{itemize}

@@

>>>>

Because each pixel in the image can be processed independently, the \gls{MVDR} beamformer is well suited for parallel acceleration. A technology of particular interest here are \glspl{GPU}~\cite{Nilsen2009}, which tailored for image processing, available off-the-shelf at reasonable prices, and are found in most desktop computers already. The \gls{GPU} is a hardware architecture comprised of several hundred computing cores, each running a thread that executes a copy of a common function called a kernel.  However, there are some inherent drawbacks with such a design: Each core is left with a rather limited amount of local memory, and distributing data to the computing cores and collecting the result is a processes with some inherent latency.

@@

To test the runtime speed of our \gls{GPU} implementation of \gls{MVDR}, we have compared it to an optimized single thread C and Matlab implementation (\Fig{benchmarks}). Processing times for a 55\,kpixels sectorscan test image are shown for varying subarray lengths $L$ and varying amounts of temporal averaging $N_\text{avg}$. The test system was a quad-core Intel Xeon E5620, with 64Gb of \gls{RAM}, and an nVidia Quadro 6000 card. This is a \gls{GPU} with 448 \gls{CUDA} cores and 6Gb of onboard \gls{RAM}, capable of roughly 1Tflops.

@@

As Figure \ref{benchmarks} illustrates, using a \gls{GPU} improves matters. If we look at the total runtimes for a subarray size of $L=16$ and $2N_\text{avg}+1=5$ pixels of temporal averaging, the \gls{GPU} implementation is able to process the test image in roughly 40\,ms. This translates to a processing capacity of 1.1\,Mpixels/s. For the same scenario the C implementation clocked in at roughly 3\,s and Matlab at 17\,s, which corresponds to a speedup factor of 75 and 435, respectively.

Interesting to note is also that solving $\eRi\1$ remains the main bottleneck in the \gls{GPU} implementation, and it gets even worse for larger $L$'s. However, this is a general Gauss Jordan solver which does not exploit symmetry properties of $\eR$, so creating e.g.\ a batch based Cholesky or UHDU solver should improve the runtimes further. Both the C and Matlab implementation in this benchmark take this symmetry into account, which partially explains why the complexity curve of these implementation have a different curve than the \gls{GPU} implementation.

% HUGIN: 1.8m/s - range 250m \\
% fs = 100kHz*30\%*$\frac{4}{3}$ = 40kHz \\
% $N_{range px} = \frac{2\,250m}{1500m/s}40kHz = 13.3kpx$ \\
% $\times 32$ beams = 426kpx \\\\
% 
% PRF = $\frac{1500m/s}{2 250m} = 3Hz$ \\
% TP = 215px 3Hz = 1.28Mpx/s \\
% 4 arrays: 1.28Mpx/s * 4 = 5.12Mpx/s
% 
% 
% \begin{itemize}
% \item Need for speed: HUGIN 4 banks of 32 elements, can be processed faster than the ping repetition rate, with margins to spare.
% \end{itemize}

@@

The \gls{MVDR} beamformer is an algorithm well suited for implementation on a \gls{GPU}. Each pixel in the image can be processed independently, and the steps within each pixel can also
